Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-25 paper/main.tex
populated all sections

\input{preamble}
\title{Accelerating Optimal Power Flow: Condensed-Space Interior-Point Methods and Automatic Differentiation on GPUs}

@@

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle

While graphics processing units (GPUs) have showcased impressive capabilities in various computing domains, their utilization in large-scale constrained nonlinear optimization regimes, such as alternating current (AC) optimal power flow (OPF) problems, has been somewhat limited. This limitation stems from the challenges associated with parallel factorization of indefinite sparse matrices commonly encountered within constrained optimization algorithms (Anitescu et. al. 2021). Although GPU computation can accelerate various other parts of the optimization process, including automatic differentiation (AD) and sparse matrix-vector multiplications, the slow data transfer between host and device memory hinders the ad-hoc implementation of GPU accelerations. To fully leverage the capabilities of modern GPU hardware, it is essential to implement a comprehensive computational framework on the GPU that incorporates AD, linear algebra, and optimization while minimizing data transfers to and from host memory.

This paper presents a comprehensive computational framework and the associated software implementations for solving AC OPF problems on GPUs. Our approach utilizes the following techniques: (i) condensed-space interior-point methods (IPMs) with an inequality relaxation strategy, (ii) sparse matrix factorization with a fixed pivot sequence, and (iii) a single-instruction, multiple-data (SIMD) abstraction of nonlinear programs. Specifically, our method relaxes power flow equality constraints by allowing small violations, which enables expressing the Karush-Kuhn-Tucker (KKT) system entirely in the primal space through the condensation procedure. Although this strategy is not new (see Nocedal and Wright, 2006), it has traditionally been considered less efficient than the standard full-space method due to increased fill-in in the sparse factorization. However, when implemented on GPUs, it offers the key advantage of guaranteeing positive definiteness in the condensed KKT system through the application of standard regularization techniques, which in turn, allows for the utilization of linear solvers with a fixed numerical pivot sequence (so-called refactorization). An efficient implementation of the sparse refactorization is available as part of the CUDA library, facilitating the implementation of efficient KKT system solutions on GPUs. Although this method is susceptible to ill-conditioning, our results demonstrate that the solver is robust enough to solve problems with a relative accuracy of $10^{-6}$.

Furthermore, by leveraging the SIMD abstraction of nonlinear programs, which preserves the parallelizable structure in the model, the model functions and derivative evaluations can be parallelized, thereby facilitating evaluations on the GPU. We demonstrate that the AC power flow model is particularly well-suited for this abstraction as it involves repetitive expressions for each type of component in the model (e.g., buses, lines, generators), and the number of computational patterns does not increase with the network's size. This structure has been effectively utilized by Gravity (Hijazi et. al. 2018), demonstrating significant acceleration in model evaluations.

The paper will present comprehensive numerical benchmark results to demonstrate the efficiency of our proposed approach. Our method is implemented in our packages, a nonlinear optimization solver MadNLP.jl and an automatic differentiation tool SIMDiff.jl, with the solution of the KKT system being carried out using the external CUSOLVER library. We compare our method against the standard CPU approach (W\"{a}chter and Beigler, 2006) and the recently developed reduced-space interior-point method on GPUs (Pacaud et. al., 2022) using the standard data available in pglib-opf. Preliminary results indicate that our proposed framework has the potential to accelerate the solution of AC OPF problems up to moderate tolerances ($10^{-6}$) by an order of magnitude compared to existing tools such as Ipopt (interfaced with MATPOWER or PowerModels.jl), especially for large-scale instances. We will conclude with a discussion on the future extensibility of our method for more complex optimization tasks, such as multi-period, security-constrained, and joint transmission-distribution optimization.

% \bibliographystyle{plain}
% \bibliography{main}

>>>>

% \input{preamble}
\title{Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods 
}

@@

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for efficiently
  solving AC optimal power flow (OPF) problems using GPUs. Although GPUs
  have demonstrated remarkable performance in various computing domains,
  their application in AC OPF has been limited due to challenges such as
  serial computations for automatic differentiation (AD) and sparse matrix
  factorization. To address these obstacles, we propose two strategies.
  First, we utilize a single-instruction, multiple-data (SIMD)
  abstraction of nonlinear programs (NLP). This approach allows us to
  specify model equations while preserving their parallelizable
  structure. By implementing AD on GPUs
  efficiently, we can fully leverage the computational power of these
  devices.  Second, we employ a Condensed-Space Interior-Point Method
  (IPM) with inequality relaxation. This technique involves relaxing
  equality constraints to inequalities and condensing the
  Karush-Kuhn-Tucker system to the primal space. As a result, we enable
  efficient sparse matrix factorization on GPUs without requiring a
  numerical pivoting procedure.  By combining these strategies, we
  achieve efficient AC OPF problem solutions on GPUs, keeping the
  problem data residing in the device memory and performing the majority
  of computations on the GPUs. Comprehensive numerical benchmark results
  showcase the substantial computational advantage of our approaches for
  solving AC OPF problems up to moderate precision on NVIDIA
  GPUs. Remarkably, our method achieves an order of magnitude speedup
  compared to state-of-the-art tools on CPUs, such as JuMP.jl, AMPL, and
  Ipopt. The paper concludes by discussing the future extensibility of
  our method.
\end{abstract}

\section{Introduction}

While graphics processing units (GPUs) have demonstrated remarkable
capabilities in various computing domains, their adoption in
large-scale constrained nonlinear optimization regimes, such as
alternating current (AC) optimal power flow (OPF) problems, has faced
some limitations. One of the primary challenges arises from the
automatic differentiation (AD) of sparse model equations and the parallel
factorization of indefinite sparse matrices commonly encountered
within constrained optimization algorithms
\cite{anitescu2021targeting}. While GPU computation can trivially
accelerate several parts of the optimization process, especially
various internal computations within the optimization solver, the
sluggish data transfer between host and device memory hampers the
ad-hoc implementation of GPU accelerations (Fig. \ref{fig:simd}). To
fully leverage the potential that modern GPU hardware has to offer, it
becomes imperative to have a comprehensive computational framework for
optimization on GPUs so that AD, linear algebra
operations, and optimization can be performed entirely on the GPU.
Specifically, for best performance, both the problem data and the
solver's intermediate computational data must be exclusively resident
within the device memory, with the majority of operations executed on
the GPU.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};

    \node[align=center] at (1.25, 3.75) {Data};
    \node[align=center] at (6.75, 3.75) {Data};

    % Host Memory
    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};

    % Device Memory
    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};

    % Arrow with Dashed Line
    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {

      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
\end{figure}

This paper presents our approach to implement a comprehensive
computational framework for solving large-scale 
OPF problems on NVIDIA GPUs, along with the associated software
implementations: SIMDiff.jl, an AD tool, and
MadNLP.jl, a nonlinear optimizaiton solver. Our approach incorporates
two novel strategies: (i) a single-instruction, multiple-data (SIMD)
abstraction of nonlinear programs (NLPs), enabling streamlined
parallel AD on GPUs, and (ii) a condensed-space
interior-point method (IPM) with an inequality relaxation strategy,
which facilitates the use of highly efficient {\it refactorization}
routines for sparse matrix factorization with fixed pivot sequences.

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results show that AD often
constitutes more than half of the total solver time when using
off-the-shelf AD implementations like JuMP.jl or AMPL. Instead of the
general-purpose tools, our method leverages a specialized AD
implementation based on the SIMD abstraction of NLPs. This abstraction
alllows to preserve the parallelizable structure within the model
equations, facilitating efficient derivative evaluations on the
GPU. The AC power flow model is particularly well-suited for this
abstraction as it involves repetitive expressions for each component
type (e.g., buses, lines, generators), and the number of computational
patterns does not increase with the network's size. Numerical results
reported in this paper demonstrate that our proposed strategies can
achieve over 20 times speedup. In comparison to general AD implementations
on CPUs (such as AMPL and JuMP.jl), our GPU-based method is
approximately 50 times faster in the largest instances.

Linear algebra operations, especially sparse indefinite matrix
factorization, are typically the bottleneck in nonlinear optimization
solvers. Parallelizing this operation is challenging due to the need
for numerical pivoting, which ensures the numerical
stability. However, when the matrix can be factorized without
numerical pivoting, a significant part of the operation can be
parallelized, and the numerical factorization can be efficiently
performed on GPUs. We develop a condensed-space IPM strategy that
allows the use of sparse matrix factorization routines without
numerical pivoting. This strategy relaxes equality constraints by
permitting small violations, which enables expressing the
Karush-Kuhn-Tucker (KKT) system entirely in the primal space
through the condensation procedure. Although this strategy is not new
\cite{nocedal2006numerical}, it has traditionally been considered less
efficient than the standard full-space method due to increased nonzero
entries in the KKT system. However, when implemented on GPUs, it
offers the key advantage of ensuring positive definiteness in the
condensed KKT system through the application of standard
regularization techniques. This, in turn, allows for the utilization
of linear solvers with a fixed numerical pivot sequence (known as
refactorization). An efficient implementation of sparse
refactorization is available as part of the CUDA library, enabling the
implementation of efficient KKT system solutions on GPUs. Although
this method is susceptible to numerical stability issues due to
increased condition number in the KKT system, our results demonstrate
that the solver is robust enough to solve problems with a relative
accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.

We present numerical benchmark results to showcase the efficiency of
our proposed approach, utilizing two of our packages: MadNLP.jl and
SIMDiff.jl. The solution of the KKT system is performed using the
external cuSOLVER library. To assess the performance of our method, we
compare it against standard CPU approaches using the widely used data
available in pglib-opf \cite{wachter2006implementation}.  Our
benchmark results demonstrate that our proposed computational
framework has significant potential for accelerating the solution of
AC OPF problems, particularly for large-scale instances, especially
when a moderate tolerance (e.g., $10^{-4}$) is
considered. Specifically, when running on NVIDIA GPUs, our method
achieves a notable 4x speedup compared to our solver running on CPU,
in the case of the largest instance. Moreover for the same instance,
our approach surpasses the performance of existing tools (such as
Ipopt interfaced with JuMP.jl) by an order of magnitude.  This finding
underscores the importance of harnessing the computational power of
GPUs as a key enabler for tackling the challenges posed by large-scale
OPF problems.

\paragraph*{Organization}
The paper is organized as follows. In the remainder of the current
section, we introduce the mathematical notation. In Section
\ref{sec:prelim}, we provide general preliminary knowledge on
numerical optimization and GPU computing. In Section \ref{sec:simd},
we present the SIMD abstraction strategy for large-scale nonlinear
programs and their advantages in terms of implementing parallel
AD. Section \ref{sec:ipm} presents the
optimization algorithm under study, the condensed-space interior point
method with inequality relaxation strategy. Section \ref{sec:num}
presents the numerical results, comparing our approach with other
state-of-the-art solution methods on GPUs. Finally, conclusions and
future outlooks are given in Section \ref{sec:conc}.

\paragraph*{Notation}
We denote the set of real numbers and the set of integers by
$\mathbb{R}$ and $\mathbb{I}$. We let $[M]:=\{1,2,\cdots,M\}$. We let
$[v_i]_{i\in[M]}:=[v_1;v_2;\cdots,v_M]$.  A vector of ones with an
appropriate size is denoted by $\boldsymbol{1}$. We use the following
convention for any symbol $x$: $X:=\diag(x)$.

\section{Preliminaries}\label{sec:prelim}
This section covers three essential background topics: numerical
optimization, AD, and GPU computing.

\subsection{Numerical Optimization}\label{sec:numopt}
We consider NLPs of the form:
\begin{subequations}\label{eqn:cpt}
  \begin{align}
    \min_{x^\flat \leq x \leq x^\sharp}\;
    & f(x)\\
    \st\;
    & g(x) =0. \label{eqn:cpt-con}
  \end{align} 
\end{subequations}
Numerous solution algorithms have been developed in the NLP
literature.  These methods can be broadly classified into active-set
methods and interior-point methods \cite{nocedal2006numerical}.
Active-set methods aim to find the set of active consraints associated
with the optimal solution in a combinatorial manner, while
interior-point methods replace inequality constraints with smooth
barrier functions, enabling the use of Newton-type second-order
algorithms. Interior-point methods are known to be more scalable for
problems with a large number of constraints and suitable for
parallelization, thanks to the fixed sparsity pattern of the KKT
matrix. We adopt the interior-point approach and develop our
optimizationmethods on GPUs.

In terms of practical computations, three key components play
vital roles: derivative evalutions (often provided by the AD
capabilities of the algebraic modeling languages), linear algebra
operations, and internal computations within the solver. Notably, most
of the computational efforts are delegated to the external linear
solver and AD library. The optimization solver
orchestrates the operation of these tools to drive the solution
iterate towards the stationary point of the optimization problem, but
computations performed within the solver are generally much cheapr and
simpler compared to the operations in AD and linear solver packages.

Since the successful implementation of the open-source interior point
solver in Ipopt, many subsequent implementations of nonlinear
optimizaiton solvers
\cite{chiang2014structured,rodriguez2023scalable,shin2021graph} have
been based on Ipopt \cite{wachter2006implementation}. We also use
Ipopt as our main reference for the IPM implementation. Below, we
outline the overall computational procedure employed wihtin nonlinear
optimization frameworks:

1) Given the current primal-dual iterate $(x^{(\ell)},y^{(\ell)},
z^{\flat(\ell)},z^{\sharp(\ell)})$, the AD package computes:
$\nabla_x f(x^{(\ell)})$, $\nabla_x g(x^{(\ell)})$, $\nabla^2_{xx}
\mathcal{L}(x^{(\ell)},y^{(\ell)},z^{\flat(\ell)},z^{\sharp(\ell)})$, where
$\mathcal{L}(x^{\flat},y^{\flat},z^{\flat},z^{\sharp}):=f(x) - y^\top
g(x) - z^\flat (x-x^\flat) - z^\sharp (x^\sharp-x)$.

2) The following sparse indefinite system (known as the KKT system) is
solved using sparse indefinite factorization (typically, via sparse
LBL$^\top$ factorization) and triangular solve routines.
\begin{align}\label{eqn:kkt-indefinite}
  &\begin{bmatrix}
    W^{(\ell)}  + \Sigma^{(\ell)} + \delta^{(\ell)}_w I& A^{(\ell)\top}\\
    A^{(\ell)} & -\delta_c I\\
  \end{bmatrix}
  \begin{bmatrix}
    \Delta x\\
    \Delta y\\
  \end{bmatrix}=
  \begin{bmatrix}
    f - \mu X^{-1} \boldsymbol{1}\\
    \Delta y\\
  \end{bmatrix}
\end{align}
Here, $W^{(\ell)}:=\nabla^{2}_{xx}\mathcal{L}(x^{(\ell)},y^{(\ell)})$,
$A^{(\ell)}:= \nabla_xg(x^{(\ell)})$, and $\delta_w, \delta_c>0$ are
the regularization parameters determined based on the inertia
correction procedure.

3) The optimization solver employs the filter line search procedure to
determine the step size. While alternative approaches, such as
trust-region-type methods and merit function-based criteria, can be
used for determining the step size, the filter line search method is
most commonly used due to its implementation in Ipopt.

\subsection{AD}
Efficiently evaluating the derivatives of $f(\cdot)$, $g(\cdot)$, and
$\mathcal{L}(\cdot,\cdot)$ is crucial for the efficient solution of
optimization problems.

Numerical differentiation of computer programs can be achieved through
three different methods: (i) finite difference method, (ii) symbolic
differentiation, and (iii) AD. The finite difference method suffers
from numerical rounding errors, and its computational complexity grows
unfavorably with respect to the number of function arguments, making
it less preferable unless no other alternatives are
available. Symbolic differentiation uses computer algebra systems to
obtain symbolic expressions of first or higher-order
derivatives. While this method can differentiate functions up to high
numerical precision, it suffers from "expression swelling" effect and
struggles to compute the derivatives of long nested expressions in a
computationally efficient way.

In contrast, AD differentiates computer
programs by leveraging the computation graph. By inspecting the
computation graph and applying chain rules, AD
can efficiently and accurately evaluate derivatives. This approach has
become the dominant paradigm for derivative computation within the
scientific computing domain, including nonlinear optimization and
machine learning. For large-scale optimization, such as AC OPFs, AD
tools are often implemented as part of domain-specific modeling
languages, such as JuMP, CasADi, and AMPL (optimization) and Torch and
Flux (machine learning).

There are two alternative ways of propagating derivatives through the
recursive application of chain rules: (i) forward-mode and (ii)
reverse-mode, which operate in opposite directions (from leaves to
root and from root to leaves, respectively). Reverse-mode automatic
differentiation, also known as an adjoint method, has proven to be
particularly effective for dealing with the function expressions in
large-scale optimization problems.

The Julia Language, our language of choice, provides convenient ways
of implementing highly efficient automatic
differentiation. \textit{Any Julia function}, including commonly used
operations such as addition, multiplication, trigonometric functions,
exponential functions, and others, can be straightforwardly extended
through the use of multiple dispatch paradigm, allowing functions to
be dynamically dispatched based on the run-time type. Several
implementations of AD are available in Julia, including
ReverseDiff.jl, ForwardDiff.jl, Zygote.jl, and JuMP.jl. While these
tools are general and useful for various applications, they are not
optimized for evaluating derivatives of AC OPF problems, which contains
an obvious parallelizable structure that can be exploited.

\subsection{GPU Computing}\label{sec:gpu}
With the increasing prevalence of GPU in various scientific computing
tasks, there has been growing interest in leveraging these parallel
systems to efficiently solve large-scale OPF problems. Some recent
concurrent approaches include HyKKT, a hybrid (direct-iterative) KKT
system solver, \cite{regev2023hykkt} and HiOP, a HPC solver for
nonlinear optimization problems, \cite{hiop_techrep}. However,
adapting the IPM to GPUs presents challenges due to the fundamental
differences between GPU and CPU programming paradigms. While CPUs
execute a sequence of instructions on a single input (single
instruction, single data, or SISD, in Flynn's taxonomy), GPUs run the
same instruction simultaneously on hundreds of threads using the SIMD
paradigm (see Fig. \ref{fig:simd}). The SIMD parallelism works well
for algorithms that can be decomposed into simple instructions running
entirely in parallel, but not all algorithms seamlessly fit this
paradigm. For example, branching in the control flow can hinder
lockstep execution across multiple threads, and in turn, prevents
efficient implementations on GPUs. On the other hand, when the
algorithm's structure allows for efficient parallelization, the SIMD
parallelism in GPUs becomes highly effective, and oftentimes enables
orders of magnitudes speedup.

We highlight that the following, arguably quite common, computational
patterns are particularly effective when implemented on GPUs:
\begin{align}
  y&\leftarrow \left[g(x; q_j)\right]_{j\in [J]}\tag{Pattern 1}\label{eqn:pattern-1}\\
  y&\leftarrow y + \sum_{k\in [K]} h(x;s_k)\tag{Pattern 2}\label{eqn:pattern-2}\\
  o&\leftarrow \sum_{i\in [I]} f(x; p_i)\tag{Pattern 3}\label{eqn:pattern-3}.
\end{align}
Here, it is assumed that $f:\mathbb{R}^{n_x}\times
\mathbb{R}^{n_{p}}\rightarrow \mathbb{R}$, $g:\mathbb{R}^{n_x}\times
\mathbb{R}^{n_{s}}\rightarrow \mathbb{R}$, and
$h:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{p}}\rightarrow
\mathbb{R}^{n_y}$ are simple instructions that require only a handful
of operations. \ref{eqn:pattern-1} is typically most effective on
GPUs, where each available thread can operate independently without
needing to simultaneously manipulate the same data reference in device
memory. \ref{eqn:pattern-2} and \ref{eqn:pattern-3} are more
challenging to implement, as they require simultaneous access to the
same device memory location, but they can be parallelized effectively
by using buffers.

Many of the operations required in AD of sparse
physical models, as well as the application of optimization
algorithms, are based on the computational patterns mentioned
above. For example, an OPF model can be implemented
with 14 different computational patterns, all of which fall within the
aforementioned categories. Additionally, the computation within
optimization solvers, such as forming the left-hand-side for the KKT
systems, computing the $\|\cdot\|_\infty$ norm of the constraint
violation, and assembling the condensed KKT system, can be carried out
by using these computational patterns. The only exception is the
factorization of the sparse KKT matrix, which requires more sophisticated
implementations.

Implementing kernel functions for the above computational patterns in
the Julia Language is straightforward, as it provides excellent
high-level interfaces for array and kernel programming for GPU
arrays. The code can even be device-agnostic, thanks to the portable
programming capability brought by KernelAbstractions.jl. All
of the AD and optimization capabilities in our
tools, MadNLP.jl and SIMDiff.jl, are implemented in Julia Language,
utilizing its kernel and array programming capabilities.

\begin{figure}[t]
  \centering 
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]
    \node[draw,align=left] (I) at (0,2.75) {
      $\bullet$ Instruction $g(\cdot,\cdot)$\\
      $\bullet$ Pointer to array $x$
    };
    \node[draw] at (-4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (A\x) at (0,-\x/2) {$s_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (0,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (C\x) at (0,-\x/2) {$y_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \foreach\x in {1,2,3,4,5,7}{
      \draw[->] (A\x.east) -- (B\x.west);
      \draw[->] (B\x.east) -- (C\x.west);
      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
    }
    \node at (-4,-2.5) {Input Data};
    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
    \node at (4,-2.5) {Output Data};
  \end{tikzpicture}
  \label{fig:simd}
  \caption{A schematic description of SIMD parallelism in GPU computing}
\end{figure}

\section{SIMD Abstraction of NLPs}\label{sec:simd}

This section describes our implementation of SIMD abstraction and
sparse AD of the model equations. The abstraction and AD are
implemented as part of our algebriac modeling language SIMDiff.jl.

The SIMD abstraction under consideration is as follows:
\begin{subequations}\label{eqn:prob}
  \begin{align}
    \min_{x^L\leq x \leq x^U}& \sum_{\ell\in[L]}\sum_{i\in [I_k]} f^{(\ell)}(x; p^{(\ell)}_i)\\
    \st\; &\forall m\in[M]:\\
                             &\begin{aligned}[t]
                               g^{(m)}_\flat\leq &\left[g^{(m)}(x; q_j)\right]_{j\in [J]} \\
                               &+\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) \leq g^{(m)}_\sharp,
                             \end{aligned}
  \end{align}
\end{subequations}
where $f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
to the first argument, ${{p^{(k)}i}{i\in [N_k]}}{k\in[K]}$,
${{q^{(k)}{i}}{i\in [M\ell]}}{\ell\in[L]}$, and
${{s^{(k)}{j}}{j\in[J\ell]}}_{\ell\in[L]}$ are problem data, which can
either be discrete or continuous. We assume that our functions
$f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ can be expressed in the form of computational
graphs of moderate length; that is, the resulting Lagrangian Hessian
and Jacobian matrices are sparse. One can observe that the problem in
\eqref{eqn:prob} is expressed by a combination of the computational
patterns in Section \ref{sec:gpu}.

Our implementation of the algebraic modeling interface enforces the
user to specify the model equations in an {\tt Iterable} data type in
Julia. Julia's {\tt Iterable} data type consists of the instruction
and the data over which the instruction is executed. This allows
maintaining the NLP problem information in the form of SIMD
abstraction in \eqref{eqn:prob}.

Now, we discuss the key benefits of the SIMD abstraction.

\paragraph{Parallel AD Implementation}
The main reason for introducing a distinctive modeling abstraction in
\eqref{eqn:prob} is because without the abstraction, it is difficult
for the AD package to detect the parallelizable structure within the
model. Many of the physics-based models, such as AC OPF, have
repetitive structure. One of the manifestations of it is
that the mathematical statement of the model is concise, even if the
model itself contains millions or even billions of variables and
constraints. Specifically, it suffices to use 15 computational
patterns to fully specify the model. These patterns arise from (1)
generation cost, (2) reference bus voltage angle constraint, (3-6)
active and reactive power flow (to and from), (7) voltage angle
difference constraint, (8-9) apparent power flow limits (to and from),
(10-11) power balance equations, (12-13) generators' contributions to
the power balance equations, and (14-15) in/out flows contributions to
the power balance equations. However, such repetitive structure is not
well exploited during AD in the standard
nonlinear optimization modelign tools. By preserving the repetitive
structures in the model, one can facilitate the implementation of
parallelized derivative computations.

SIMDiff.jl implements the standard reverse-mode automatic
differentiation. Using the multiple dispatch feature of Julia,
SIMDiff.jl can generate highly efficient derivative computation code,
specifically compiled for each computational pattern in the model. Furthermore,
the derivative evaluation code can straightforwardly extended so that
it can run over the data on acceraltor devices. In particular,
the derivative computation code is obtained as a type-stable function,
which can be run on devices through array and kernel programming.

\paragraph{Memory Efficiency}
The memory footprint of storing model
information can be significantly reduced with the SIMD abstraction. As
opposed to storing the symbolic expressions and sparsity patterns for
each constraint and objective term, which can potentially be more than
millions, we only store the symbolic expression and sparsity patterns
for each computational pattern. This allows drastically reducing the
memory requirements. Thanks to this, our implementation has a
significantly shorter model creation time and smaller memory footprint
compared to the existing sparse AD tools.

\paragraph{Efficient Symbolic Analysis}
The symbolic analysis is essential in sparse automatic
differentiation. This step is needed for determining the sparsity
pattern of the evaluated derivatives. In sparse AD, the initial
symbolic analysis of nonlinear expressions can be expensive, as the
number of objective terms and constraints can be more than
millions. However, often times those analyses are applied for the same
computational patterns, and the time and memory spent for symbolic
analysis can be significantly reduced if the SIMD-compatible
computational patterns are present in the model. SIMDiff.jl exploits
the SIMD abstraction of the model equations to save the computational
cost spent for symbolic analysis, and this allows for efficient
evaluation of derivatives without the use of the expensive symbolic
anlaysis nor coloring procedures.

\section{Condensed-Space Interior Point Methods with Inequality Relaxation Strategy}\label{sec:ipm}
We present the condensed-space interior point method within the
context of the general NLP formulation in \eqref{eqn:cpt}. Our method
has two key differences from standard interior point method
implementations: (i) the use of inequality relaxation and (ii) the
condensed treatment of the KKT system.

\subsection{Inequality Relaxation}

At the beginning of the algorithm, we apply inequality relaxation to replace the equality constraints in \eqref{eqn:cpt} with inequalities by introducing slack variables $s\in\mathbb{R}^{m}$:
\begin{align}\label{eqn:relax}
  g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
\end{align}
where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are chosen to be close to zero.
This relaxed problem can be stated as follows:
\begin{align}\label{eqn:relaxed} 
  \begin{aligned}[t]
    \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
    &  f(x)\\
    \st\;
    & g(x) - s = 0.
  \end{aligned}
\end{align}

In our implementation, we set $s^{\flat},s^\sharp$ as
$-\epsilon_{\text{tol}}$ and $+\epsilon_{\text{tol}}$, respectively,
where $\epsilon_{\text{tol}}>0$ is a user-specified relative tolerance
of the interior point method (IPM). This type of relaxation is
commonly used in practical IPM implementations; for example, in Ipopt,
the solver relaxes the bounds and inequality constraints by
$O(\epsilon_{\text{tol}})$ to prevent an empty interior of the
feasible set (see \cite[Section 3.5]{wachter2006implementation}).
For condensed-space IPM, we cannot maintain the same level of precision
due to the increased condition number of the KKT system.
We have found that setting $\epsilon_{\text{tol}}$ to be
$\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$ ensures numerical
stability while achieving satisfactory convergence. Thus, our solver
sets the tolerance to $10^{-4}$ by default when using condensed IPM.

\subsection{Barrier Subproblem}

The barrier subproblem is considered as follows:
\begin{subequations}\label{eqn:barrier}
  \begin{align}
    \min_{x}\;
    &
      \begin{aligned}[t]
        &f(x) - \mu\boldsymbol{1}^\top \log (x- x^\flat) -\mu\boldsymbol{1}^\top \log (x^\sharp -x)\\
        &- \mu\boldsymbol{1}^\top \log (s- s^\flat) -\mu\boldsymbol{1}^\top \log (s^\sharp -s)
      \end{aligned}
          \label{eqn:barrier-obj}\\
    \st\;
    &g(x) = 0. \label{eqn:barrier-con}
  \end{align}
\end{subequations}
Here, $\mu>0$ is the barrier parameter. The smooth log-barrier
function is employed to avoid handling inequalities in a combinatorial
fashion (as in active set methods). A superlinear local convergence to
the first-order stationary point can be achieved by repeatedly
applying Newton's step to the KKT conditions of \eqref{eqn:barrier}
with $\mu=\mu^{(\ell)}$ and a decreasing sequence of
${\mu^{(\ell)}}_{\ell=0,1,\cdots}$. A global convergence can be
achieved by employing a certain globalization strategy (either based
on a merit function or filter method) \cite{nocedal2006numerical}.

\subsection{Newton's Step Computation}
The Newton step direction can be computed by considering the first-order optimality conditions for the barrier subproblem in \eqref{eqn:barrier}:
\begin{align}\label{eqn:first} 
  \begin{aligned}[t]
    \nabla_{x} f(x^{(\ell)}) - \nabla_{x}g(x^{(\ell)})^\top y^{(\ell)}  - z_x^\flat  + z_x^\sharp = 0\;&\\
    \begin{aligned}
      - z_s^\flat  + z_s^\sharp  &= 0\\
      Z^\flat_x (x^{(\ell)}-x^\flat) - \mu\boldsymbol{1} &= 0\\
      Z^\flat_s (s^{(\ell)}-s^\flat) - \mu\boldsymbol{1}&= 0,
    \end{aligned}
    \quad
    \begin{aligned}
      g(x^{(\ell)}) &= 0\\
      Z^\sharp_x (x^\sharp-x^{(\ell)}) - \mu\boldsymbol{1}&= 0\\
      Z^\sharp_s (s^\sharp-s^{(\ell)}) - \mu\boldsymbol{1}&= 0,
    \end{aligned}&
  \end{aligned}
\end{align}
where $y\in\mathbb{R}^{m}$, $z_x^\flat,z_x^\sharp\in\mathbb{R}^{n}$,
and $z_s^\flat,z_s^\sharp\in\mathbb{R}^{m}$ are Lagrange multipliers
associated with the equality and bound constraints in
\eqref{eqn:relaxed}. The Newton step for solving the nonlinear
equation in \eqref{eqn:first} can be computed by solving the following
KKT system in \eqref{eqn:very-long-eqn}.
\begin{figure*}[h!]
  % \hline
  \begin{align}\label{eqn:very-long-eqn}
    \begin{bmatrix}
      W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
      & \delta^{(\ell)}_w I & -I&&&-I & I\\
      A^{(\ell)}& -I & -\delta_c I\\
      Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
      -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
      &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
      &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
    \end{bmatrix}
    \begin{bmatrix}
      \Delta x \\
      \Delta s \\
      \Delta y \\
      \Delta z_x^\flat \\
      \Delta z_x^\sharp \\
      \Delta z_s^\flat \\
      \Delta z_s^\sharp \\
    \end{bmatrix} =
    \begin{bmatrix}
      p_{x }\\
      p_{s }\\
      p_{y }\\
      p_{z_x^\flat }\\
      p_{z_x^\sharp }\\
      p_{z_s^\flat }\\
      p_{z_s^\sharp }\\
    \end{bmatrix}
  \end{align}
  % \hline
\end{figure*}
Here, $p_x,\cdots p_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
\eqref{eqn:first}. In what follows, we shall drop the superscript
$(\cdot)^{(\ell)}$ for concise notation.
